Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-05-30 paper/manuscript.tex
Adding reference to ABFE part

\paragraph{Have you considered stereoisomers and enantiomers?}
Congeneric series can contain stereoisomers or enantiomers which can bind very differently, resulting in large errors if treated incorrectly. For racemates, the relative abundance of each stereoisomer is normally not known, however when there is one stereocenter in the molecule it can often be fair to assume that the substance is a 50\%:50\% mixture of the two. If one enantiomer is active (eutomer) and the other is inactive (distomer), this corresponds to the effective concentration of your bioactive molecule being half of what is expected in the assay, which corresponds to a ~0.2 pIC50 loss in potency relative to the pure eutomer being tested, which could be captured within the assoicated error calculations. If possible, both enantiomers could be considered computationally, however if a series conatins multiple stereocenters, or multiple compounds within the series have stereocenters, this may be computationally restrictive. If compounds contain multiple undefined stereoecenters, the error and complication of understanding predictions may become prohibitive.Therefore, the experimental activity associated with just one stereoisomer/enantiomer is more uncertain. If enumerating unknown stereocenters is not tractable, modeling using just the conformation that best fits the active site is reasonable, however  this clearly this introduces potential for larger errors compared to experiment. Despite this, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer. Additionally, unexpected changes in what stereoisomer binds experimentally, if they occur, could pose significant challenges for modelling efforts.

\paragraph{Are there cofactors expected to be present?}
Crystallographic structures can contain cofactors, which may be close to the ligand series of interest, and understanding if they are required or not for simulation can be an interdisciplinary effort between the modeller, crystallographer and bioligists to establish if they are a crystallographic artifact, or if they are likely to be present in the assay. In addition to both the crystallographic and assay conditions, what is most relevant _in vivo_ might be a third, equally difficult consideration. Following this, if a cofactor is considered to be included in the calculations, it is worth considering the appropriate forcefield for the cofactors. Cofactors, in particular metal ions can be challenging\cite{mobley2017predicting}. If the congeneric series of ligands is expected to interact with the cofactors differently, such as metal ions in different charge states, or to displace the cofactor for some of the molecules then this is likely beyond current modelling techniques. If there is still uncertainty around cofactor binding, running free energy calculations for a set of molecules both with and without the cofactor and comparing the relative rank-ordering of molecules could provide a sanity check.

@@

Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed. One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation. Few descriptions of the approach can be found for such approaches like the one suggested by Baumann et al.~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute} for seperated topology approach, finding , or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other. 
However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets. \cite{jespers2019qligfep, Rieder2022Lev}

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address thisusing a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.

@@

\paragraph{Several choices of restraints are possible.}
In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.

>>>>

\paragraph{Have you considered stereoisomers and enantiomers?}
Congeneric series can contain stereoisomers or enantiomers which can bind very differently, resulting in large errors if treated incorrectly. For racemates, the relative abundance of each stereoisomer is normally not known. Therefore, the experimental activity associated with just one stereoisomer/enantiomer is more uncertain. However, the modeling typically uses just the bioactive conformation that best fits the active site. Clearly this introduces potential for larger errors compared to experiment. Nevertheless, if all compounds in the congeneric series are racemic, originating from similar synthetic procedures with an expected similar abundance of stereoisomers, then the differences may cancel and the trend in calculated and observed binding energies may be robust. Despite this, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer. Additionally, unexpected changes in what stereoisomer binds experimentally, if they occur, could pose significant challenges for modelling efforts.

@@

Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed. One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation . This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}). Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other. 
However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets. \cite{jespers2019qligfep, Rieder2022Lev}

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.

@@

\paragraph{Several choices of restraints are possible.}\label{ABFErest}
In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.
